Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-07-05 write-ups/FSE/paper.tex
fixed one thing with vspace
hopefully one of the final passes through the paper

\begin{document}

@@

@@

\myparagraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many interesting data-parallel algorithms are \emph{irregular}: the
amount of work to be processed is unknown ahead of time and may be
determined by the computation itself.  There is growing interest in
accelerating such algorithms on
GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}. \TSAdded{Such
  algorithms include octree partitioning, which has applications in
  particle simulations; and graph traversals, which have applications
  in speech decoding.}

@@

%\vspace{-1mm}
\myparagraph{A degree of fairness: occupancy-bound execution} The current GPU programming
models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
HSA~\cite{HSAprogramming11}, specify almost no guarantees regarding
scheduling of workgroups, and current GPU schedulers are unfair in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.\footnote{In practice,
  depending on the kernel, multiple workgroups might map to the same
  compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle more workgroups being
launched than there are compute units is via an \emph{occupancy-bound}
execution
model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where,
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it finishes execution.
This simple model is easy to implement
efficiently, and experiments suggest that
it is widely employed by today's
GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.

@@

%\vspace{-1mm}
\myparagraph{Resistance to occupancy-bound execution}
Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multitasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
tasks, such as graphics rendering.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to
power down some compute units if the battery level is low.
Implementing energy
throttling would require being able to suspend the execution of a
workgroup, powering down the compute unit to which the workgroup was
bound, something that the occupancy-bound execution model does not allow.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that GPU vendors (1) will
not commit to the occupancy-bound execution model they currently
implement, for the above reasons, yet (2) will not guarantee fair
scheduling \TSAdded{using preemption. This is due to the high runtime
  cost of preempting workgroups on GPUs, which requires managing
  thread local state (e.g. registers, program location) for all
  threads in a workgroup (up to 1024 on Nvidia GPUs), as well as
  \emph{shared memory}, the workgroup local cache (up to 64 KB on
  Nvidia GPUs)}.  Vendors instead wish to retain the essence of the
  simple occupancy-bound model, supporting preemption only in key
  special cases.

For example, Nvidia's latest architecture (Pascal) is documented to
support instruction-level preemption~\cite{PascalWhitepaper}. Despite
this, Nvidia's most recent compute capability does not guarantee
forward progress for CUDA programmers (evidence that vendors will not
yet commit to providing such guarantees).  Indeed, on a GTX Titan X
(Pascal) we still observe starvation: we found that a global barrier
can be executed successfully with 56 workgroups, but deadlocks with 57
workgroups.  This is an example of a vendor supporting preemption, yet
not guaranteeing fair scheduling.

%\vspace{-1mm}
\myparagraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms
demand fair scheduling,
% of workgroups,
but for good reasons
GPU vendors will not commit even to the guarantees of the
occupancy-bound execution model.
%
We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.

@@

\TSAdded{Cooperative kernels allow for \emph{cooperative multitasking}
  (see \mysec{\ref{sec:relatedwork}}), a multitasking approach used
  historically when preemption was not available or, in the case of
  GPUs, too costly (see previous paragraph). Our approach avoids the
  cost of preemptive multitasking as the state of a workgroup killed
  via $\offerkill$ does not have to be saved. Previous cooperative
  multitasking systems often provide \emph{yield} based semantics,
  where a processing unit would temporarily give up its hardware
  resource. We intentionally deviate from this design as in the case
  of a global barrier (a common idiom in the applications we study),
  adopting yield semantics would force the cooperative kernel to block
  \emph{completely} in the case where a single workgroup yields,
  stalling the kernel until the given workgroup resumes.  Instead, our
  $\offerkill$ allows a kernel to make progress with a smaller number
  of workgroups, with workgroups potentially joining again later via
  $\offerfork$.}

@@

\item The model is backwards-compatible: if $\offerkill$ and $\offerfork$ are ignored, a cooperative kernel will behave
  exactly as a regular kernel does on current GPUs.

@@

The threads of a kernel are divided into \emph{workgroups}.
%, each consisting of a fixed number of threads.
Functions
$\getlocalid$ and $\getgroupid$ return a thread's local id within
its workgroup and the workgroup id.
%
%\footnote{For simplicity we assume 1D kernels, which captures all the
%irregular algorithms we know of, and write
%e.g.\ $\getglobalid$ for $\getglobalid(0)$.
%}
%
The number
of threads per workgroup and number of workgroups are obtained via
$\getlocalsize$ and $\getnumgroups$.
%A thread's local and global
%id are related by $\getglobalid = \getgroupid \times
%\getlocalsize + \getlocalid$.  The total number of threads executing the
%kernel is given by $\getglobalsize = \getlocalsize \times \getnumgroups$.

Execution of the threads in a workgroup can be synchronised via a
workgroup barrier. This barrier is a \emph{workgroup-level} function:
it can only appear in conditional code if all threads in a workgroup
that reach the barrier agree on the guards of all enclosing
conditional statements.  A \emph{global} barrier (synchronising all
threads of a kernel) is \emph{not} provided as a primitive.

\myparagraph{Memory spaces and memory model} A kernel has access to
four memory spaces.  \emph{Shared virtual memory} (SVM) is accessible
to all device threads and the host concurrently.  \emph{Global} memory is
shared among all device threads.  Each workgroup has a
portion of \emph{local} memory for fast intra-workgroup communication.
Every thread has a portion of very fast \emph{private} memory for
function-local variables.

@@

\myparagraph{Execution model}
OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
  conforming implementation may choose to serialize the workgroups
%so
%  a correct algorithm cannot assume that workgroups will execute in
%  parallel.
\dots There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.
%
HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... (if) A
  comes after B in work-group flattened ID order}. This is not sufficient to support blocking algorithms that use
mutexes and inter-workgroup barriers, both of which require \emph{symmetric} communication between
threads.

@@

\myparagraph{Work stealing}
%
Work stealing enables dynamic balancing of tasks across
processing units. It is useful when the number of tasks to be
processed is dynamic, due to one task creating an arbitrary number of
new tasks.  Work stealing has been explored in the context of
GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}. Each workgroup has a
queue from which it obtains tasks to process, and to which
it stores new tasks. If its queue is empty, a workgroup
tries to \emph{steal} a task from another queue.

@@

If a task was obtained, then the workgroup processes it
(line~\ref{line:wksteal:processtask}), which may lead to new tasks
being created and pushed to the workgroup's queue. The kernel presents
two opportunities for spin-waiting: spinning to obtain a mutex, and
spinning in the main kernel loop to obtain a task.

@@

\myparagraph{Graph traversal} \myfiglong\ref{fig:graphsearch} illustrates a frontier-based graph traversal algorithm; such algorithms have
been shown to execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}.
The kernel is
given three arguments in global memory: a graph structure, and two
arrays of graph nodes. Initially, $\keyword{n0}$ contains the
starting nodes to process. Private variable $\keyword{level}$ records the current frontier level, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ point to
distinct arrays recording the nodes to be processed during the current and next frontier, respectively.

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between
threads through \emph{stride} based processing.
%
In this case, the stride is the total number of threads, obtained via
$\getglobalsize$.  Each thread processes nodes with indices
\code{$\keyword{tid}$+$\keyword{stride}$*0},
\code{$\keyword{tid}$+$\keyword{stride}$*1},
\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
A thread calls $\keyword{process\_node}$ to process a node given the current level, with nodes to be processed next frontier being pushed to $\keyword{out\_nodes}$. After processing the frontier, the threads swap their
node array pointers (line~\ref{line:graph:swap}).

@@

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global barrier
can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}, based
on CPU barrier designs~\cite[ch. 17]{HS08}.  These barriers employ
spinning to ensure threads wait at the barrier until all threads have
arrived, which requires fair scheduling between workgroups for the
barrier to operate correctly.

The mutexes and barriers used by these two examples appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units.  This is due to the fairness
of the occupancy-bound execution model that current GPUs have been
shown, experimentally, to provide.  But, as discussed in
\mysec\ref{sec:intro}, this model is not endorsed by language
standards or vendor implementations, and is unlikely to be respected
in the future. \TSAdded{Without a fair schedule, the mutexes and
  barriers, which use spin-waiting in their implementation, are not
  guaranteed to be satisfied. For example, if a workgroup is not
  fairly scheduled after obtaining a mutex, there is no guarantee that
  the mutex will be released. Thus the kernel may hang while other
  workgroups wait indefinitely to obtain the mutex. }
%

@@

We present our cooperative kernels programming model as an extension
to OpenCL.  However, the cooperative kernels concept is more general,
and could be applied to extend other GPU programming models, such as
CUDA and HSA.  We describe the semantics of the model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}) and outline
important nonfunctional properties that the model requires to work
successfully (\mysec\ref{sec:nonfunctional}).
%and explain that the
%model is backwards-compatible
%(\mysec\ref{sec:backwardscompatibility}).
%

@@

As with a regular OpenCL kernel (see \mysec\ref{sec:opencl}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  Unlike in a regular kernel, the parameters to a cooperative kernel are immutable (though pointer
parameters can refer to mutable data).

@@

%\vspace{-1mm}
\myparagraph{Transmitted variables}
%
A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  Annotating
a variable $v$ with $\transmit$ means that when a workgroup 
uses $\offerfork$ to spawn new
workgroups, the workgroup should transmit its
current value for $v$ to the threads of the new workgroups.
We detail the semantics for this when we
describe $\offerfork$ below.

%\vspace{-1mm}
\myparagraph{Active workgroups}
%
If the host application launches a cooperative kernel requesting $N$
workgroups, this indicates that the kernel should be executed with a
maximum of $N$ workgroups, and that as many workgroups as possible, up
to this limit, are desired.  However, the scheduler may initially
schedule fewer than $N$ workgroups, and as explained below the number
of workgroups that execute the cooperative kernel can change during
the lifetime of the kernel.

@@

Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
$\offerkill$, is a workgroup-level function---it must be encountered
uniformly by all threads in a workgroup (see
\mysec\ref{sec:opencl}).

@@

\myfiglong{\ref{fig:overview}} illustrates the behaviour of
$\offerfork$. After the graphics task finishes executing, there is
space for two workgroups to join the computation of the cooperative
kernel. Workgroup $0$ calls $\offerfork$, spawning the two new
workgroups with ids $2$ and $3$ and broadcasting the transmit
variables to them. Workgroups $2$ and $3$ join the computation where
workgroup $0$ called $\offerfork$.

%We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
%having all new threads obtain this local state portion from thread 0,
%as well as our reasons for selecting which variables to transmit via
%annotations rather than transmitting the entire thread state.

@@

%\vspace{-1mm}
\myparagraph{Global barriers}
%
Because workgroups of a cooperative kernel are fairly scheduled, so a
global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
and $\resizingglobalbarrier$.

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented by adapting an inter-workgroup barrier
design, e.g.~\cite{XF10}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
the OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

@@

\subsection{Programming With Cooperative Kernels}\label{sec:programmingguidelines}

@@

\myparagraph{A changing workgroup count} Unlike in regular OpenCL, the
value returned by $\getnumgroups$ is not fixed during the lifetime of
a cooperative kernel: it corresponds to the active group count $M$,
which changes as workgroups execute $\offerkill$, and $\offerfork$.
The value returned by $\getglobalsize$ is similarly subject to change.
A cooperative kernel must thus be written in a manner that is robust
to changes in the values returned by these functions.

@@

\myparagraph{Adapting work stealing}
%
In this example there is no state to transmit since a computation is
entirely parameterised by a task, which is retrieved from a queue
located in global memory. With respect to \myfig~\ref{fig:workstealing},
we add $\offerfork$ and $\offerkill$ calls at the start of the main loop
(below line~\ref{line:wksteal:mainloop}) to let a workgroup offer itself
to be killed or forked, respectively, before it processes a task.  Note
that a workgroup may be killed even if its associated task queue is not
empty, since remaining tasks will be stolen by other workgroups. In
addition, since $\offerfork$ may be the entry point of a workgroup, the
queue id must now be computed after it, so we move
line~\ref{line:wksteal:getgroupid} to be placed just before
line~\ref{line:wksteal:poporsteal}. In particular, the queue id cannot
be transmitted since we want a newly spawned workgroup to read its own
queue and not the one of the forking workgroup.

@@

Our vision is that support for cooperative kernels will be integrated
in the runtimes of future GPU implementations of OpenCL, with driver
support for our new primitives.  To experiment with our ideas on
current GPUs, we have developed a prototype that mocks up the required
runtime support via a \emph{megakernel}, and exploits the
occupancy-bound execution model that these GPUs provide to ensure fair
scheduling between workgroups.  We emphasise that an aim of
cooperative kernels is to \emph{avoid} depending on the
occupancy-bound model.  Our prototype exploits this model simply to
allow us to experiment with current GPUs whose proprietary drivers we
cannot change.  We describe the megakernel approach
(\mysec\ref{sec:megakernel}) and detail various aspects of the
scheduler component of our implementation
(\mysec\ref{sec:schedulerimpl}).

\subsection{The megakernel mock up}\label{sec:megakernel}

>>>>

\usepackage{flushend}
\begin{document}

@@

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10010940.10010941.10010949.10010957.10010959</concept_id>
<concept_desc>Software and its engineering~Multiprocessing / multiprogramming / multitasking</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011039.10011311</concept_id>
<concept_desc>Software and its engineering~Semantics</concept_desc>
<concept_significance>100</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010371.10010387.10010389</concept_id>
<concept_desc>Computing methodologies~Graphics processors</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Multiprocessing / multiprogramming / multitasking}
\ccsdesc[100]{Software and its engineering~Semantics}
\ccsdesc[300]{Computing methodologies~Graphics processors}

\keywords{GPU, cooperative multitasking, irregular parallelism}

@@

\myparagraph{The Needs of Irregular Data-parallel Algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many interesting data-parallel algorithms are \emph{irregular}: the
amount of work to be processed is unknown ahead of time and may be
determined during computation.  There is growing interest in
accelerating such algorithms on
GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.

@@

%\vspace{-1mm}
\myparagraph{A Degree of Fairness: Occupancy-bound Execution} The current GPU programming
models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
HSA~\cite{HSAprogramming11}, specify almost no guarantees regarding
scheduling of workgroups, and current GPU schedulers are unfair in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.\footnote{In practice,
  depending on the kernel, multiple workgroups might map to the same
  compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle more workgroups being
launched than there are compute units is via an \emph{occupancy-bound}
execution
model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where,
once a workgroup has commenced execution on a compute unit (it has
become \emph{occupant}), the workgroup has exclusive access to the
compute unit until it finishes execution.
Experiments suggest that this model
is widely employed by today's
GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.

@@

%\vspace{-1mm}
\myparagraph{Resistance to Occupancy-bound Execution}
Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multitasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
tasks (e.g.\ graphics rendering).
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is an important concern for
battery-powered
devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the
future, it will be desirable for a mobile GPU driver to power down
some compute units, and hence suspend the execution of the occupant
workgroup, if the battery level is low.
%Implementing energy
%throttling would require being able to suspend the execution of a
%workgroup, powering down the compute unit to which the workgroup was
%bound, something that the occupancy-bound execution model does not allow.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}),
is that GPU vendors (1) will not commit to the occupancy-bound
execution model they currently implement, for the above reasons, yet
(2) will not guarantee fair scheduling \TSAdded{using preemption. This
  is due to the high runtime cost of preempting workgroups, which
  requires managing thread local state (e.g. registers, program
  location) for all workgroup threads (up to 1024 on Nvidia
  GPUs), as well as \emph{shared memory}, the workgroup local cache
  (up to 64 KB on Nvidia GPUs)}.  Vendors instead wish to retain the
essence of the simple occupancy-bound model, supporting preemption
only in key special cases.

For example, Nvidia's Pascal architecture supports
preemption~\cite{PascalWhitepaper}. Despite this, the device still
does not guarantee forward progress. Indeed, on a GTX Titan X (Pascal)
we still observe starvation: we found that a global barrier can be
executed successfully with 56 workgroups, but deadlocks with 57
workgroups.  Thus, although preemption is supported, there is not fair
scheduling.

%\vspace{-1mm}
\myparagraph{Our Proposal: Cooperative Kernels}
%
To summarise: blocking algorithms
demand fair scheduling,
% of workgroups,
but for good reasons
GPU vendors will not commit to the guarantees of the
occupancy-bound execution model.
%
We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.

@@

\TSAdded{Cooperative kernels allow for \emph{cooperative multitasking}
  (see \mysec{\ref{sec:relatedwork}}), a multitasking approach used
  historically when preemption was not available or, in the case of
  GPUs, too costly. Our approach avoids the cost of arbitrary
  preemption as the state of a workgroup killed via $\offerkill$ does
  not have to be saved. Previous cooperative multitasking systems have
  provided \emph{yield} semantics, where a processing unit would
  temporarily give up its hardware resource. We deviate from this
  design as, in the case of a global barrier, adopting yield would
  force the cooperative kernel to block \emph{completely} in the case
  where a single workgroup yields, stalling the kernel until the given
  workgroup resumes.  Instead, our $\offerkill$ allows a kernel to
  make progress with a smaller number of workgroups, with workgroups
  potentially joining again later via $\offerfork$.}

@@

\item The model is backwards-compatible: $\offerkill$ and $\offerfork$ may be ignored, and a cooperative kernel will behave
  exactly as a regular kernel does on current GPUs.

@@

The threads of a kernel are divided into \emph{workgroups}.
%, each consisting of a fixed number of threads.
Functions
$\getlocalid$ and $\getgroupid$ return a thread's local id within
its workgroup and the workgroup id.
%
%\footnote{For simplicity we assume 1D kernels, which captures all the
%irregular algorithms we know of, and write
%e.g.\ $\getglobalid$ for $\getglobalid(0)$.
%}
%
The number
of threads per workgroup and number of workgroups are obtained via
$\getlocalsize$ and $\getnumgroups$.
%A thread's local and global
%id are related by $\getglobalid = \getgroupid \times
%\getlocalsize + \getlocalid$.  The total number of threads executing the
%kernel is given by $\getglobalsize = \getlocalsize \times \getnumgroups$.
%
Execution of the threads in a workgroup can be synchronised via a
workgroup barrier. 
%This barrier is a \emph{workgroup-level} function:
%it can only appear in conditional code if all threads in a workgroup
%that reach the barrier agree on the guards of all enclosing
%conditional statements.  
A \emph{global} barrier (synchronising all
threads of a kernel) is \emph{not} provided as a primitive.

\myparagraph{Memory Spaces and Memory Model} A kernel has access to
four memory spaces.  \emph{Shared virtual memory} (SVM) is accessible
to all device threads and the host concurrently.  \emph{Global} memory is
shared among all device threads.  Each workgroup has a
portion of \emph{local} memory for fast intra-workgroup communication.
Every thread has a portion of very fast \emph{private} memory for
function-local variables.

@@

\myparagraph{Execution Model}
OpenCL~\cite[p.\ 31]{opencl2Spec} and CUDA~\cite{cuda-75} specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel. 
%stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
%  conforming implementation may choose to serialize the workgroups
%so
%  a correct algorithm cannot assume that workgroups will execute in
%  parallel.
%\dots There is no safe and portable way to synchronize across
%  the independent execution of workgroups since once in the work-pool,
%  they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.
%
HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... (if) A
  comes after B in work-group flattened ID order}. This is not sufficient to support blocking algorithms that use
mutexes and inter-workgroup barriers, both of which require \emph{symmetric} communication between
threads.

@@

\myparagraph{Work Stealing}
%
Work stealing enables dynamic balancing of tasks across
processing units. It is useful when the number of tasks to be
processed is dynamic, due to one task creating an arbitrary number of
new tasks.  Work stealing has been explored in the context of
GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}. Each workgroup has a
queue from which it obtains tasks to process, and to which
it stores new tasks. If its queue is empty, a workgroup
tries to \emph{steal} a task from another queue.

@@

If a task was obtained, then the workgroup processes it
(line~\ref{line:wksteal:processtask}), which may lead to new tasks
being created and pushed to the workgroup's queue. The kernel presents
two opportunities for spin-waiting: spinning to obtain a mutex, and
spinning in the main kernel loop to obtain a task. \TSAdded{Without fair
scheduling, threads waiting for the mutex might spin indefinitely,
causing the application to hang.}

@@

\myparagraph{Graph Traversal} \myfiglong\ref{fig:graphsearch} illustrates a frontier-based graph traversal algorithm; such algorithms have
been shown to execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}.
The kernel is
given three arguments in global memory: a graph structure, and two
arrays of graph nodes. Initially, $\keyword{n0}$ contains the
starting nodes to process. Private variable $\keyword{level}$ records the current frontier level, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ point to
distinct arrays recording the nodes to be processed during the current and next frontier, respectively.

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between
threads through \emph{stride} based processing.
%
In this case, the stride is the total number of threads, obtained via
$\getglobalsize$.  
%Each thread processes nodes with indices
%\code{$\keyword{tid}$+$\keyword{stride}$*0},
%\code{$\keyword{tid}$+$\keyword{stride}$*1},
%\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
A thread calls $\keyword{process\_node}$ to process a node given the current level, with nodes to be processed next frontier being pushed to $\keyword{out\_nodes}$. After processing the frontier, the threads swap their
node array pointers (line~\ref{line:graph:swap}).

@@

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global barrier
can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}, based
on CPU barrier designs~\cite[ch. 17]{HS08}.  These barriers employ
spinning to ensure threads wait at the barrier until all threads have
arrived, which requires fair scheduling between workgroups for the
barrier to operate correctly. \TSAdded{Without fair scheduling, the
  barrier threads may wait indefinitely at the barrier, causing the
  application to hang.}

The mutexes and barriers used by these two examples appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units.  This is due to the fairness
of the occupancy-bound execution model that current GPUs have been
shown, experimentally, to provide.  But, as discussed in
\mysec\ref{sec:intro}, this model is not endorsed by language
standards or vendor implementations, and may not be respected
in the future. 
%

@@

We present our cooperative kernels programming model as an extension
to OpenCL.  
%However, the cooperative kernels concept is more general,
%and could be applied to extend other GPU programming models, such as
%CUDA and HSA.  
We describe the semantics of the model
(\mysec\ref{sec:semantics}), use our motivating examples to discuss
programmability (\mysec\ref{sec:programmingguidelines}) and outline
important nonfunctional properties that the model requires to work
successfully (\mysec\ref{sec:nonfunctional}).
%and explain that the
%model is backwards-compatible
%(\mysec\ref{sec:backwardscompatibility}).
%

@@

As with a regular OpenCL kernel, a cooperative kernel is launched by
the host application, passing parameters to the kernel and specifying
a desired number of threads and workgroups. Unlike in a regular
kernel, the parameters to a cooperative kernel are immutable (though
pointer parameters can refer to mutable data).

@@

%\vspace{-1mm}
\myparagraph{Transmitted Variables}
%
A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  Annotating
a variable $v$ with $\transmit$ means that when a workgroup 
uses $\offerfork$ to spawn new
workgroups, the workgroup should transmit its
current value for $v$ to the threads of the new workgroups.
We detail the semantics for this when we
describe $\offerfork$ below.

%\vspace{-1mm}
\myparagraph{Active Workgroups}
%
If the host application launches a cooperative kernel requesting $N$
workgroups, this indicates that the kernel should be executed with a
maximum of $N$ workgroups, and that as many workgroups as possible, up
to this limit, are desired.  However, the scheduler may initially
schedule fewer than $N$ workgroups, and as explained below the number
of workgroups that execute the cooperative kernel can change during
the lifetime of the kernel.

@@

Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
$\offerkill$, is a workgroup-level function---it must be encountered
uniformly by all threads in a workgroup.

@@

\myfiglong{\ref{fig:overview}} illustrates the behaviour of
$\offerfork$. After the graphics task finishes executing, workgroup
$0$ calls $\offerfork$, spawning the two new workgroups with ids $2$
and $3$. Workgroups $2$ and $3$ join the computation where workgroup
$0$ called $\offerfork$.
%
%We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
%having all new threads obtain this local state portion from thread 0,
%as well as our reasons for selecting which variables to transmit via
%annotations rather than transmitting the entire thread state.
%

@@

%\vspace{-1mm}
\myparagraph{Global Barriers}
%
Because workgroups of a cooperative kernel are fairly scheduled, so a
global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
and $\resizingglobalbarrier$.

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented by adapting an inter-workgroup barrier
design, e.g.~\cite{XF10}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
the OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  
%However, implementing such a barrier is
%involved, hence why we include this function as a primitive.

@@

\subsection{Programming with Cooperative Kernels}\label{sec:programmingguidelines}

@@

\myparagraph{A Changing Workgroup Count} Unlike in regular OpenCL, the
value returned by $\getnumgroups$ is not fixed during the lifetime of
a cooperative kernel: it corresponds to the active group count $M$,
which changes as workgroups execute $\offerkill$, and $\offerfork$.
The value returned by $\getglobalsize$ is similarly subject to change.
A cooperative kernel must thus be written in a manner that is robust
to changes in the values returned by these functions.

@@

\myparagraph{Adapting Work Stealing}
%
In this example there is no state to transmit since a computation is
entirely parameterised by a task, which is retrieved from a queue
located in global memory. With respect to \myfig~\ref{fig:workstealing},
we add $\offerfork$ and $\offerkill$ calls at the start of the main loop
(below line~\ref{line:wksteal:mainloop}) to let a workgroup offer itself
to be killed or forked, respectively, before it processes a task.  Note
that a workgroup may be killed even if its associated task queue is not
empty, since remaining tasks will be stolen by other workgroups. In
addition, since $\offerfork$ may be the entry point of a workgroup, the
queue id must now be computed after it, so we move
line~\ref{line:wksteal:getgroupid} to be placed just before
line~\ref{line:wksteal:poporsteal}. In particular, the queue id cannot
be transmitted since we want a newly spawned workgroup to read its own
queue and not the one of the forking workgroup.

@@

Our vision is that cooperative kernel support will be integrated
in the runtimes of future GPU implementations of OpenCL, with driver
support for our new primitives.  To experiment with our ideas on
current GPUs, we have developed a prototype that mocks up the required
runtime support via a \emph{megakernel}, and exploits the
occupancy-bound execution model that these GPUs provide to ensure fair
scheduling between workgroups.  We emphasise that an aim of
cooperative kernels is to \emph{avoid} depending on the
occupancy-bound model.  Our prototype exploits this model simply to
allow us to experiment with current GPUs whose proprietary drivers we
cannot change.  We describe the megakernel approach
(\mysec\ref{sec:megakernel}) and detail various aspects of the
scheduler component of our implementation
(\mysec\ref{sec:schedulerimpl}).

\subsection{The Megakernel Mock Up}\label{sec:megakernel}
